---
output: github_document
---

# DataRaft

For Positron, see the [IDE integration guide](https://github.com/dataraft-r/dataraft/blob/feat/positron-ide/examples/positron-ide/README.md)
and [phase checklist](https://github.com/dataraft-r/dataraft/blob/feat/positron-ide/docs/plans/positron-ide.md). IDE metadata and the separate
extension are optional; existing R workflows remain usable without them.

[![R-CMD-check](https://github.com/dataraft-r/dataraft/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/dataraft-r/dataraft/actions/workflows/R-CMD-check.yaml)
[![Coverage](https://github.com/dataraft-r/dataraft/actions/workflows/coverage.yaml/badge.svg)](https://github.com/dataraft-r/dataraft/actions/workflows/coverage.yaml)
[![Lifecycle: experimental](https://img.shields.io/badge/lifecycle-experimental-orange.svg)](https://lifecycle.r-lib.org/articles/stages.html#experimental)

Keep R-based reports reproducible with checked data and pinned input versions.
DataRaft is the contract, quality and reproducibility layer for teams whose
business logic lives in R. Use your existing database or an open lakehouse,
or start with compiler-free RDS releases.
A changed column type or a negative amount blocks a delivery before its target
is written. The result retains the checks and offending rows for diagnosis.

See the [one-page cheatsheet](CHEATSHEET.md), [roadmap](ROADMAP.md) and
[security policy](SECURITY.md). Numeric coverage is published per component and
commit in the [Coverage workflow summaries](https://github.com/dataraft-r/dataraft/actions/workflows/coverage.yaml).

DataRaft is under development. The hardening focus is `dataraft.core`,
`dataraft.lake` and `dataraft.adapters`. The dbt and metrics integrations
are experimental and can change independently. Metrics provides governed,
frozen metric results; it is not a general-purpose semantic modeling engine.
No stable public release is claimed by the development compatibility lock.

Start without a database: `library(dataraft); demo <- dr_demo()`.
Inspect `dataraft.core::dr_quality_rows(demo$blocked)` and `dr_collect(demo$passed)`.
See the [insurance walkthrough](https://github.com/dataraft-r/dataraft/blob/main/examples/insurance-delivery.md) and the
[family compatibility policy](FAMILY_COMPATIBILITY.md).
The [review implementation status](https://github.com/dataraft-r/dataraft/blob/main/docs/review-hardening.md) records changes
and the remaining scope boundaries.

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

```{r failure}
library(dataraft)

orders <- dr_product("orders") |>
  dr_add_contract(c(id = "integer", amount = "numeric")) |>
  dr_add_quality(~ amount >= 0)

orders <- orders |>
  dataraft.core::dr_add_recipe(dataraft.core::dr_recipe() |> dataraft.core::dr_step_mutate(amount = round(amount, 2)))

bad_delivery <- data.frame(id = 1:3, amount = c(25, -75, 50))
checked <- dr_run(orders, data = bad_delivery, write = FALSE, stop_on_failure = FALSE)
checked$status
dataraft.core::dr_quality_rows(checked)
```

Correct the delivery and reuse the same product:

```{r success}
next_delivery <- data.frame(id = 1:3, amount = c(25, 75, 50))
result <- dr_run(orders, data = next_delivery, write = FALSE)
dr_collect(result)
```

A **product** contains the delivery, transformations, contract, checks and target.
`dr_run(write = FALSE)` checks it without invoking framework writers;
`dr_run()` executes its configured destination. `dr_collect()` retrieves checked
output. Source reads and user callbacks still run in both modes.

## Start with the essentials

Use `dr_product()`, `dr_contract()`, `dataraft.core::dr_add_source()`, `dr_add_quality()`,
`dr_quality()`, `dr_set_target()`, `dr_run()` and `dr_collect()` first.
Ordinary dplyr verbs and optional recipes prepare the delivery. Use
`dataraft.core::dr_contract_policy()` and `dataraft.core::dr_contract_meta()` when more policy or metadata is
needed. Compose directly on products; check without writing with `dr_run(write = FALSE)`.

Without a declared contract, inferred schema checks and validation reports say
`unvalidated`. Successful execution alone is not a validation guarantee.
See the [current API guide](https://github.com/dataraft-r/dataraft/blob/main/docs/plans/current-api.md)
for composition rules and execution guarantees.

Start with [why DataRaft](https://dataraft-r.github.io/dataraft/articles/why-dataraft.html)
and the [guided introduction](https://dataraft-r.github.io/dataraft/articles/get-started.html).

## Packages

| Package | Responsibility |
|---|---|
| `dataraft` | Metapackage and shared introduction |
| `dataraft.core` | Products, contracts, recipes, workflows and quality |
| `dataraft.lake` | Lake storage, releases, coordinated publication and recovery |
| `dataraft.adapters` | Database, API, RDS, Parquet and pins adapters; targets integration |
| `dataraft.dbt` | dbt execution and artifacts |
| `dataraft.metrics` | Metrics and frozen report evidence |
| `dataraft.ide` | Optional Positron metadata bridge and contract editor requests |

Use `library(dataraft.core)` for in-memory work without the extensions. Each
extension can be installed with its declared dependencies. The
metapackage re-exports the public family API; its own R code contains no engine.

Install the development version:

```r
install.packages("pak")
pak::pak("dataraft-r/dataraft")
```

## When to use it

Use DataRaft when repeated deliveries need a reusable preparation definition,
quality gates and an inspectable execution result. Use pointblank for standalone
data validation, targets for pipeline scheduling, pins for board-based object
storage and dbt for SQL model development. DataRaft adapters connect these tools
to a checked delivery workflow.

See the [introduction](https://dataraft-r.github.io/dataraft/articles/get-started.html),
[integration guide](https://dataraft-r.github.io/dataraft/articles/integrations.html) and
[guarantees and limits](https://dataraft-r.github.io/dataraft/articles/guarantees.html).


For persistence without an optional storage engine, use
`dr_publish(flow, data = delivery, to = dataraft.adapters::dr_target_rds("data/orders"))`.
The result's `outputs$version` pins that delivery for `dataraft.adapters::dr_source_rds()`.
After a failed pipe, `dr_last_failure()` retrieves the failed result for diagnosis.
