---
title: "Products, recipes and workflows"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Products, recipes and workflows}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", error = FALSE)
```

```{r failure}
library(dataraft)

orders <- dr_product("orders") |>
  dr_add_contract(c(id = "integer", amount = "numeric")) |>
  dr_add_quality(~ amount >= 0)

flow <- dr_workflow() |>
  dr_add_product(orders) |>
  dr_add_recipe(dr_recipe() |> dr_step_mutate(amount = round(amount, 2)))

bad_delivery <- data.frame(id = 1:3, amount = c(25, -75, 50))
checked <- dr_trial(flow, data = bad_delivery)
checked$status
dr_quality_rows(checked)
```

Correct the delivery and reuse the same workflow:

```{r success}
next_delivery <- data.frame(id = 1:3, amount = c(25, 75, 50))
result <- dr_trial(flow, data = next_delivery)
dr_collect(result)
```

A **product** defines identity, contract and quality requirements. A **recipe**
defines preparation in step order. A **workflow** binds these definitions to
sources and a target. Definitions do no I/O. `dr_trial()` checks without writing;
`dr_run()` executes the configured target. `dr_collect()` retrieves the output.


Use `dr_plan()` to inspect a definition before reading sources. Contract failures
and quality failures are available in the result; `dr_quality_rows()` exposes
retained failing rows. Set a destination with `dr_set_target()` and execute with
`dr_run()` once the trial passes. Quality gates run before the destination writer.

Without an explicit contract, an exploratory result is `unvalidated`. You can
collect its retained data, but it is not a passed validation and no target is
written. Add a declared contract, as above, before publishing. Inferred column
types describe the current delivery; they do not independently check it.

A passed trial checks its retained delivery. It does not check destination
permissions, capacity or a future publication conflict, and a later run can read
a different external delivery. Treat trial and publication outcomes separately.

For a reusable contract, declare its schema first and add policy and ownership
in small steps. These constructors perform no I/O:

```{r contract-composition}
contract <- dr_contract("orders", columns = c(id = "integer", amount = "numeric")) |>
  dr_contract_policy(allow_extra = FALSE) |>
  dr_contract_meta(owner = "Analytics")
```

When revising an already registered contract, use `dr_update_contract()` and an
explicit new version; construction helpers do not replace version review.

Recipes own preparation in modular workflows. Products added with
`dr_add_product()` must not carry embedded transformations. Recipe steps run in
the order added, followed by output contract checks and quality rules.

## Inspect a failed rule

Keep the result before collecting it. Failed predicates and errors while evaluating
predicates are different: the latter have original R conditions for local inspection.

```{r diagnose}
broken <- dr_product("orders", next_delivery) |> dr_add_quality(~ amont >= 0)
result <- dr_trial(broken)
dr_quality_report(result)
lapply(dr_quality_errors(result), conditionMessage)
```

When catching an error from `dr_run()` or `dr_collect()`, its `result` field also
retains the diagnosis. `dr_quality_errors(error)` accepts that condition directly.
Raw exception text is not included in exported HTML or JSON quality reports.

## Save locally without DuckDB

`dr_target_rds()` writes versioned RDS releases using only R and the hard
DataRaft dependencies. Keep the returned version to reproduce a delivery.

```{r local-rds}
folder <- tempfile("dataraft-rds-")
saved <- dr_publish(flow, data = next_delivery, to = dr_target_rds(folder))
reference <- saved$outputs$version
restored <- dr_read_source(dr_source_rds(folder, version = reference))
stopifnot(sum(restored$amount) == 150)
unlink(folder, recursive = TRUE)
```

The adapter serializes local writers with a directory lock and never overwrites
an existing version. It does not provide lake transactions across tables or a
frozen-report registry. Use `dataraft.lake` when those guarantees are required.

After a failed pipe, use `result <- dr_last_failure()` to recover its checks
without repeating source reads. Clear the retained reference with
`dr_last_failure(clear = TRUE)` when you are done.

## Inspect a definition before execution

`dr_plan()` shows planned work; `dr_validate()` checks configuration without
reading sources when given a definition. `dr_inspect()` describes an object;
`dr_status()` summarizes the outcome of an execution. None substitutes for checking
a real delivery with `dr_trial()`.
