---
title: "Products, recipes and workflows"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Products, recipes and workflows}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

```{r failure}
library(dataraft)

orders <- dr_product("orders") |>
  dr_add_contract(c(id = "integer", amount = "numeric")) |>
  dr_add_quality(~ amount >= 0)

flow <- dr_workflow() |>
  dr_add_product(orders) |>
  dr_add_recipe(dr_recipe() |> dr_step_mutate(amount = round(amount, 2)))

bad_delivery <- data.frame(id = 1:3, amount = c(25, -75, 50))
checked <- dr_trial(flow, data = bad_delivery)
checked$status
dr_quality_rows(checked)
```

Correct the delivery and reuse the same workflow:

```{r success}
next_delivery <- data.frame(id = 1:3, amount = c(25, 75, 50))
result <- dr_trial(flow, data = next_delivery)
dr_collect(result)
```

A **product** defines identity, contract and quality requirements. A **recipe**
defines preparation in step order. A **workflow** binds these definitions to
sources and a target. Definitions do no I/O. `dr_trial()` checks without writing;
`dr_run()` executes the configured target. `dr_collect()` retrieves the output.


Use `dr_plan()` to inspect a definition before reading sources. Contract failures
and quality failures are available in the result; `dr_quality_rows()` exposes
retained failing rows. Set a destination with `dr_set_target()` and execute with
`dr_run()` once the trial passes. Quality gates run before the destination writer.

Recipes own preparation in modular workflows. Products added with
`dr_add_product()` must not carry embedded transformations. Recipe steps run in
the order added, followed by output contract checks and quality rules.

## Inspect a failed rule

Keep the result before collecting it. Failed predicates and errors while evaluating
predicates are different: the latter have original R conditions for local inspection.

```{r diagnose}
broken <- dr_product("orders", next_delivery) |> dr_add_quality(~ amont >= 0)
result <- dr_trial(broken)
dr_quality_report(result)
lapply(dr_quality_errors(result), conditionMessage)
```

When catching an error from `dr_run()` or `dr_collect()`, its `result` field also
retains the diagnosis. `dr_quality_errors(error)` accepts that condition directly.
Raw exception text is not included in exported HTML or JSON quality reports.

## Save locally without DuckDB

The existing pins adapter can write RDS files to a local versioned board. This
needs `pins` and `dataraft.adapters`, but does not need DuckDB. Use
`install.packages("pins")` to install the optional engine.

```{r local-pins, eval=requireNamespace("pins", quietly=TRUE)}
folder <- tempfile("dataraft-board-")
board <- pins::board_folder(folder, versioned = TRUE)
target <- dr_target_pins(board, "orders", type = "rds")
saved <- dr_publish(flow, data = next_delivery, to = target)
reference <- saved$outputs$version
restored <- dr_read_source(dr_source_pins(board, "orders", version = reference))
stopifnot(sum(restored$amount) == 150)
unlink(folder, recursive = TRUE)
```

Pins versions are board versions. They do not provide the lake's coordinated
publication, relational transactions or frozen-report registry. Use
`dataraft.lake` and `dataraft.metrics` when those guarantees are required.

## Inspect a definition before execution

`dr_plan()` shows planned work; `dr_validate()` checks configuration without
reading sources when given a definition. `dr_inspect()` describes an object;
`dr_status()` summarizes the outcome of an execution. None substitutes for checking
a real delivery with `dr_trial()`.
