---
title: "Review contracts as files and retain delivery evidence"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Review contracts as files and retain delivery evidence}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

A reproducible report needs more than a saved total: retain the input release
identifiers, the metric definition and the checks made against those inputs.
`dataraft.metrics::dr_report_release()` in dataraft.metrics freezes that evidence for lake-backed
reports. The compiler-free example below first establishes a checked, versioned
delivery; it does not claim that an RDS delivery is itself a report release.

```{r}
library(dataraft)
contract <- dr_contract("orders",
columns = c(id = "integer", amount = "numeric"),
key = "id") |>
  dataraft.core::dr_contract_meta(owner = "Reporting",
governance = list(steward = "Data operations", retention = "P7Y",
    tags = c("Finance.Reporting")),
column_metadata = list(amount = list(description = "Order value in EUR")),
producer = "Reporting")
file <- tempfile(fileext = ".json")
dataraft.adapters::dr_contract_odcs(contract, file)
reviewed <- dataraft.adapters::dr_contract_from_odcs(file)
reviewed
```

This is ODCS 3.2.0, validated against the upstream JSON Schema in CI. Exported
metadata includes roles, SLAs and classifications when supplied. Policies are
metadata, not automatic enforcement: setting retention does not schedule deletion
and identifying a steward does not grant access.

The importer executes flat R-compatible types, required columns, primary keys,
`nullValues` and `rowCount` library checks with simple numeric comparisons, and
DataRaft custom predicates. It rejects unsupported executable constraints,
relationships and engines. It does not execute SQL or R strings from contracts.
DataRaft predicates allow column names, scalar literals, arithmetic, comparisons,
logical operators, `is.na()` and literal membership via `%in%` and `c()`.
Other engines must implement that custom rule type before executing it.

```{r}
root <- tempfile("orders-")
product <- dr_product("orders", data.frame(id = 1:3, amount = c(25, -75, 50))) |>
  dr_add_contract(reviewed) |>
  dr_add_quality(~ amount >= 0, action = "quarantine", dimension = "conformity") |>
  dr_set_target(dataraft.adapters::dr_target_rds(root))
result <- dr_run(product)
dr_collect(result)
dataraft.core::dr_quarantine_rows(result)
dr_quality_report(result)
```

Quarantine is for native table row predicates. It removes all rejected rows
before publication, even if a threshold would otherwise tolerate them. Remaining
rows must still pass schema, type, key and required-column checks. Exceptions
block execution. Quarantine rows stay local; decide explicitly where sensitive
rejected rows may be stored. Model-level or aggregate checks cannot automatically
partition member tables.

For warning and blocking rules, `threshold = .02` permits up to two percent
failures. Existing `severity` and `max_failure` remain supported.

```{r}
before <- dataraft.core::dr_profile_snapshot(data.frame(id = 1:3, amount = c(25, 75, 50)), "input-v1")
after <- dataraft.core::dr_profile_snapshot(data.frame(id = 1:3, amount = c(25, NA, 50)), "input-v2")
dataraft.core::dr_profile_compare(before, after)
profile_path <- tempfile("profiles-")
profile_release <- dr_product("orders.profile", after) |>
  dr_set_target(dataraft.adapters::dr_target_rds(profile_path)) |>
  dr_run()
profile_release$outputs$version
```

Profiles carry an observation timestamp and explicit input version. Store each
profile as a versioned release and compare a chosen baseline, rather than silently
replacing it. Missingness changes are deterministic alerts, not a calibrated
statistical anomaly detector. Exact profiles can still expose sensitive counts.

```{r}
recipe <- dataraft.core::dr_recipe() |>
  dataraft.core::dr_step_mutate(net = amount / 1.19) |>
  dataraft.core::dr_step_select(id, net)
dataraft.core::dr_column_lineage(recipe, c("id", "amount"))
```

The direct-value map handles simple mutation, selection and rename. An opaque
function or dynamic selector marks the analysis incomplete. Successful single-input
runs publish complete maps through OpenLineage column facets. This does not infer
row influence, joins, historical graphs or arbitrary R side effects.

For a file-first project, place `project.yaml`, an ODCS contract and a CSV source
in one directory. `dataraft.adapters::dr_project_yaml("project.yaml") |> dr_run()` reads this manifest:

```yaml
id: orders
contract: orders.odcs.json
source: delivery.csv
target: releases/orders
```

All paths stay inside the project directory. Review the files in Git. Add custom
R transformations explicitly in trusted project code when the declarative grammar
is insufficient.

Iceberg is an experimental alternative: `dataraft.adapters::dr_source_iceberg()` reads a metadata
location with an optional snapshot ID; `dataraft.adapters::dr_target_iceberg()` writes to an already
attached DuckDB Iceberg REST catalog. Authentication and catalog attachment stay
outside definitions. The backend must support transactional candidate validation;
no fallback, retention guarantee or branch/merge behavior is implied. See
[DuckDB's Iceberg catalog documentation](https://duckdb.org/docs/current/core_extensions/iceberg/catalogs).

```{r, include=FALSE}
unlink(c(root, profile_path, file), recursive = TRUE)
```
