---
title: "Why DataRaft?"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Why DataRaft?}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

A monthly script starts small: read orders and add their amounts. Repeated
deliveries make its assumptions part of your reporting process.

```{r deliveries}
library(dataraft)
first <- data.frame(id = 1:2, amount = c(25, 75))
second <- transform(first, amount = as.character(amount))
third <- rbind(first, first[1, ])
fourth <- transform(first, amount = c(25, 80))
sum(first$amount)  # 100
sum(third$amount)  # 125: a duplicated order silently changes the total
sum(fourth$amount) # 105: a correction changes the current figure
```

The second delivery changes a numeric column to text. The third duplicates an
order. The fourth corrects a value, leaving an old report impossible to reproduce
if its source was overwritten. These are different problems: schema, identity,
and versioned evidence.

Declare the first two assumptions once and check every delivery against them.

```{r checked}
orders <- dr_product("orders") |>
  dr_add_contract(dr_contract(columns = c(id = "integer", amount = "numeric"),
key = "id") |>
  dataraft.core::dr_contract_policy(allow_extra = FALSE)) |>
  dr_add_quality(~ amount >= 0)
flow <- orders
results <- lapply(list(first, second, third, fourth), function(delivery) {
  dr_run(write = FALSE, stop_on_failure = FALSE, flow, data = delivery)
})
vapply(results, function(result) result$status, character(1))
dr_quality_report(results[[3]])
```

## A contract makes accidental changes visible

Ordinary dplyr semantics allow a new column called `amont`. A strict output
contract makes that typo visible after preparation.

```{r typo}
preparation <- dataraft.core::dr_recipe() |> dataraft.core::dr_step_mutate(amont = round(amount, 2))
unchecked <- dr_product("orders", first) |> dataraft.core::dr_add_recipe(preparation)
names(dr_collect(dr_run(write = FALSE, stop_on_failure = FALSE, unchecked)))
checked <- orders |> dataraft.core::dr_add_source(first) |> dataraft.core::dr_add_recipe(preparation)
dr_quality_report(dr_run(write = FALSE, stop_on_failure = FALSE, checked))
```

Checking alone does not preserve history. Publish accepted deliveries to
`dataraft.lake` for pinned releases, and use `dataraft.metrics` to freeze report
values with their evidence. A correction creates a new release while the earlier
report continues to identify its original inputs. The local pins example in
[the introduction](get-started.html) offers a smaller first persistence step,
with the different guarantees described there.

Use DataRaft when these requirements repeat across deliveries or teams. For one
standalone validation, pointblank may be enough; for scheduling tasks, use targets.
The component packages connect these responsibilities while keeping definitions
reusable and execution explicit.
