# Data Contracts {#sec-contracts}

A contract records what consumers may rely on. In DataRaft, the structural constructor is deliberately small: identity, columns, keys, rules, and version.

## Start with shape and identity

```{r}
library(dataraft)

policy_contract <- dr_contract(
  "policies",
  columns = c(
    policy_id = "integer",
    broker_id = "integer",
    premium = "numeric",
    status = "character"
  ),
  key = "policy_id",
  version = "1.0.0"
)

policy_contract
```

Supported audited column types include `character`, `integer`, `numeric`, `logical`, `Date`, `POSIXct`, `integer64`, and `list`.

A key is more than a useful name. It states that the declared field is part of row identity and must satisfy the key checks.

## Add business meaning separately

```{r}
policy_contract <- policy_contract |>
  dataraft.core::dr_contract_meta(
    owner = "Risk Analytics",
    description = "Policy portfolio used by the recurring reporting workflow",
    grain = "One row per policy"
  )
```

Metadata describes the product. It does not itself decide whether a delivery is accepted.

## Add validation policy separately

```{r}
policy_contract <- policy_contract |>
  dataraft.core::dr_contract_policy(
    allow_empty = FALSE,
    allow_extra = FALSE,
    constraints = list(
      premium = list(min = 0),
      status = list(enum = c("active", "lapsed", "paid_up"))
    )
  )
```

This separation keeps three questions visible:

1. What fields form the interface?
2. What do those fields mean?
3. How strict should validation be?

## Semantics before schema

A schema says that `premium` is numeric. It does not say whether the value is annual premium, monthly premium, gross premium, or net premium.

For a reusable product, establish at least four semantic facts:

- **grain**: what one row represents;
- **identity**: which fields distinguish rows;
- **meaning**: what important measures and categories represent;
- **ownership**: who can decide whether that meaning changes.

DataRaft models part of this directly through keys and contract metadata. It does not try to become a complete enterprise semantic layer. Broader metric semantics, universal join rules, and cross-tool query semantics can remain in a dedicated semantic system.

The important design rule is that schema validation should not be mistaken for complete business meaning.

## Contract evolution

Changing rows is not the same as changing the interface.

```{r}
policy_contract_v2 <- dataraft.core::dr_contract_update(
  policy_contract,
  version = "2.0.0",
  columns = c(
    policy_id = "integer",
    broker_id = "integer",
    premium = "numeric",
    status = "character",
    sales_channel = "character"
  )
)

dataraft.core::dr_contract_diff(policy_contract, policy_contract_v2)
```

A useful rule is:

- new rows with the same meaning: new run, same contract version;
- changed consumer expectation: new contract version;
- changed storage location only: usually no contract change.

Compatibility remains a consumer question. Adding a field may be harmless for one consumer and breaking for another. Renaming a key or changing row grain deserves explicit downstream review.

::: {.callout-tip title="Design guidance"}
Put business meaning in metadata, stable structural expectations in the contract, and delivery-specific outcome checks in quality rules.
:::

## Exercises

1. Add `sales_channel` to a new contract version and explain whether existing consumers need to change.
2. Change the key from `policy_id` to two columns. What statement about row identity have you changed?
3. Write one semantic statement about `premium` that cannot be represented by its R type alone.
