# Data Quality as an Execution Gate {#sec-quality}

Contracts describe the interface. Quality rules decide whether the current delivery is acceptable.

## Start from one decision

```{r}
library(dataraft)

policy_delivery <- read.csv("data/policies.csv")

policies <- dr_product("policies") |>
  dr_add_contract(dr_contract(
    "policies",
    columns = c(
      policy_id = "integer",
      broker_id = "integer",
      premium = "numeric",
      status = "character"
    ),
    key = "policy_id"
  )) |>
  dr_add_quality(dataraft.core::dr_quality_rule(
    "premium_non_negative",
    ~ premium >= 0,
    action = "block",
    threshold = 0,
    dimension = "accuracy"
  ))
```

A rule should exist because its result changes what happens next.

## Block, warn, or quarantine

The current API supports three actions:

- `block`: reject the delivery under the current rule;
- `warn`: permit the delivery but retain a warning;
- `quarantine`: remove failing rows before writing and retain the rejected rows locally on the result.

Quarantine is intentionally narrow. It is available when DataRaft can identify the failing rows of a native row predicate. It is not a generic mechanism for ignoring errors.

## Inspect a blocked result

```{r}
bad_delivery <- policy_delivery
bad_delivery$premium[3] <- -5

blocked <- dr_run(
  policies,
  data = bad_delivery,
  write = FALSE,
  stop_on_failure = FALSE
)

blocked$status
dr_quality_report(blocked)
dataraft.core::dr_quality_rows(
  blocked,
  rule = "premium_non_negative"
)
```

The report answers which checks passed or failed. The row view answers which records were involved when row evidence exists.

## Thresholds change policy

A threshold is the permitted fraction of failed test units. A zero threshold requires every evaluated unit to pass.

```{r}
warning_rule <- dataraft.core::dr_quality_rule(
  "small_negative_share",
  ~ premium >= 0,
  action = "warn",
  threshold = 0.10
)

warning_product <- dr_product("policies_warning") |>
  dr_add_quality(warning_rule)

warning_result <- dr_run(
  warning_product,
  data = bad_delivery,
  write = FALSE,
  stop_on_failure = FALSE
)

dr_quality_report(warning_result)
```

Changing a threshold can change which deliveries reach consumers. Treat it as an acceptance-policy change, not merely tuning.

## Structural checks and business checks

A missing required column means the interface is broken. A negative premium means values violate a business expectation. Both may block publication, but they point to different owners and different fixes.

This distinction improves diagnosis:

| Question | Best home |
|---|---|
| Is the required field present? | contract |
| Is the key unique and non-null? | contract |
| Is premium non-negative? | quality rule or stable constraint |
| Is a small amount of drift tolerable? | quality policy |
| Who owns the meaning? | metadata |

## Optional Pointblank execution

Pointblank is a specialized validation engine. DataRaft can use it for checks while retaining its own product-level gate and evidence model. That keeps the core workflow usable without Pointblank and lets richer validation remain an optional responsibility.

::: {.callout-important title="Failure evidence is part of the product story"}
A useful quality system explains a rejected delivery. The failed result, rule identity, affected rows, and retained evidence are as important as the passing case.
:::

## Exercises

1. Run the same defect with `block` and `warn`. Compare the result status and quality report.
2. Choose a real rule where quarantine would have a clear business meaning. Explain why dropping the row would be safer than blocking the whole delivery.
3. Change a threshold from zero to a positive value and describe the governance decision you have made.
