# Transformations and Recipes {#sec-recipes}

A recipe records ordered preparation separately from product identity, contracts, and destinations. It does not replace ordinary R transformation code. It makes a preparation sequence inspectable and reusable when that is valuable.

## Prepare the policy delivery

```{r}
library(dataraft)

preparation <- dataraft.core::dr_recipe() |>
  dataraft.core::dr_step_mutate(
    premium = round(premium, 2)
  ) |>
  dataraft.core::dr_step_filter(
    status != "cancelled"
  )

preparation
```

A recipe can be attached without executing it.

```{r}
policies <- dr_product(
  "policies",
  read.csv("data/policies.csv")
) |>
  dataraft.core::dr_add_recipe(preparation)

result <- dr_run(policies, write = FALSE)
dr_collect(result)
```

The current recipe API includes familiar mutate, filter, select, rename, arrange, summarise, and distinct steps. `dr_step_transform()` provides an escape hatch for custom functions.

## Lookups make dependencies explicit

The book's broker reference data is a meaningful dependency of the portfolio product.

```{r}
brokers <- read.csv("data/brokers.csv")

portfolio_recipe <- dataraft.core::dr_recipe() |>
  dataraft.core::dr_step_lookup(
    brokers,
    by = "broker_id",
    name = "brokers"
  )
```

The output contract belongs to the data after preparation. If the lookup adds `region`, the published interface should say whether `region` is part of the product.

## Recipes and static column lineage

For a bounded set of mutate, select, and rename operations, DataRaft can inspect direct field dependencies without running the recipe.

```{r}
lineage_recipe <- dataraft.core::dr_recipe() |>
  dataraft.core::dr_step_mutate(
    rounded_premium = round(premium, 0)
  ) |>
  dataraft.core::dr_step_select(
    policy_id,
    rounded_premium
  )

dataraft.core::dr_column_lineage(
  lineage_recipe,
  c("policy_id", "premium")
)
```

Unsupported transformations return `complete = FALSE` rather than a guessed map.

## Keep ordinary code ordinary

Use a recipe when the ordered preparation itself should be inspectable, composed, tested, or connected to lineage. Use an ordinary R function when a local transformation is clearer that way.

If a recipe grows into many transformations with several independently useful outputs, that is often a sign that the workflow contains more than one product.

::: {.callout-note title="Reproducibility boundary"}
Recipe expressions retain their R environments. Prefer explicit product sources for business data and stable constants for configuration. A recipe is inspectable preparation, not an isolation sandbox.
:::

## Exercises

1. Add a rename step and inspect the resulting column lineage.
2. Replace one declarative step with a custom function. What can static inspection no longer prove?
3. Decide whether the payment aggregation in the running example belongs inside a policy recipe or deserves a separate product at a new grain.
