# Storage and Persistence {#sec-storage}

Storage is a physical capability attached to a product. It should not be the only place where the product's meaning exists.

## Add the smallest durable target

The local RDS adapter is useful for learning because it needs no service or credentials.

```{r}
library(dataraft)

policy_delivery <- read.csv("data/policies.csv")

policies <- dr_product("policies", policy_delivery) |>
  dr_add_contract(dr_contract(
    "policies",
    columns = c(
      policy_id = "integer",
      broker_id = "integer",
      premium = "numeric",
      status = "character"
    ),
    key = "policy_id"
  )) |>
  dr_add_quality(~ premium >= 0)

release_root <- tempfile("policy-releases-")

published <- policies |>
  dr_set_target(dataraft.adapters::dr_target_rds(release_root)) |>
  dr_publish()

published$status
published$outputs
```

Each successful RDS write creates a new immutable version directory and checksum metadata. Publication uses local filesystem locking and staging rename semantics. Those are useful local guarantees, not a distributed transaction.

## Read an exact release

A versioned destination becomes more valuable when a consumer can name the exact accepted output it used.

```{r}
version <- published$outputs$version

restored <- dataraft.core::dr_read_source(
  dataraft.adapters::dr_source_rds(
    release_root,
    version
  )
)

restored
```

Leaving the source version unresolved can be useful for "latest accepted data" discovery. Pinning the version is stronger evidence for a frozen report or reproducible analysis.

## Definition version and release version are different

A contract version such as `1.0.0` identifies a definition. An RDS release such as `v000000000001` identifies one accepted physical publication.

A product can publish many releases under one unchanged contract. Conversely, a new contract version can exist before any delivery satisfies it.

Do not collapse both ideas into one generic "version."

## Other physical adapters

`dataraft.adapters` also provides database, Parquet, pins, API, metadata, and experimental Iceberg integrations. Their capabilities differ.

A database may provide transactions that a filesystem does not. A Parquet dataset has different update semantics from a mutable table. An API source has availability and replay limits that a local file does not.

The adapter boundary gives these components a common place in the workflow without pretending their guarantees are identical.

## When the lake becomes useful

`dataraft.lake` adds a registry, immutable release identity, lifecycle state, recovery, integrity verification, and lake-oriented storage choices.

A useful adoption sequence is:

```text
in memory
  -> versioned RDS
  -> governed lake when release operations require it
```

The step to a lake should be triggered by an operating requirement such as shared release history, recovery, lifecycle state, or registry-backed discovery. It should not be triggered by the desire to make a small project look more like a platform.

::: {.callout-tip title="Physical change test"}
If moving from RDS to a lake forces you to redefine product grain, keys, or business meaning, the physical layer has probably leaked too far into the logical definition.
:::

## Exercises

1. Publish the policy product twice and read both explicit versions.
2. List three guarantees the RDS adapter provides and two it does not.
3. Decide which concrete requirement would justify moving this example from RDS to a DataRaft lake.
