---
title: "Define, check and publish a product"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Define, check and publish a product}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", error = FALSE)
```

A product contains its sources, preparation, contract, checks and destination.
Start with an explicit contract: without one, inferred schema checks are labelled
`unvalidated`. A completed run then means execution succeeded, not that a business
schema was validated.

```{r define}
library(dataraft)

contract <- dr_contract("orders", columns = c(id = "integer", amount = "numeric"),
                        key = "id") |>
  dataraft.core::dr_contract_meta(owner = "Analytics", grain = "One order")
orders <- dr_product("orders", contract = contract) |>
  dataraft.core::dr_add_recipe(dataraft.core::dr_recipe() |> dataraft.core::dr_step_mutate(amount = round(amount, 2))) |>
  dr_add_quality(dr_quality(~ amount >= 0, action = "block"))
```

Use `action = "warn"` for a non-blocking check and `threshold` for an allowed
failure fraction. Quarantine removes rejected rows before writing. It does not
silently relax the rule's stored policy. Custom checks must be stable: ordinary
checks are evaluated twice during dry runs and known time/random calls are rejected.
Productive runs reuse a successful comparison for the same rule definition within
the R session; without one they verify first. Explicit
`volatile = TRUE` allows them, labels their evidence, disables release caching and
prevents that evidence from authorizing frozen report approvals.

```{r failure}
bad_delivery <- data.frame(id = 1:3, amount = c(25, -75, 50))
checked <- dr_run(orders, data = bad_delivery, write = FALSE, stop_on_failure = FALSE)
checked$status
dataraft.core::dr_quality_rows(checked)
```

Correct the delivery and reuse the product:

```{r success}
next_delivery <- data.frame(id = 1:3, amount = c(25, 75, 50))
result <- dr_run(orders, data = next_delivery, write = FALSE)
dr_collect(result)
```

`write = FALSE` uses the same checks and transformations while suppressing
framework targets, catalogs and run-evidence writes, including upstream products.
It still reads sources and executes your functions. Those functions can perform
I/O themselves. A later run reads a new delivery; this check does not reserve it.

## Save locally

```{r local-rds}
folder <- tempfile("dataraft-rds-")
saved <- dr_publish(orders, data = next_delivery, to = dataraft.adapters::dr_target_rds(folder))
reference <- saved$outputs$version
restored <- dataraft.core::dr_read_source(dataraft.adapters::dr_source_rds(folder, version = reference))
stopifnot(sum(restored$amount) == 150)
unlink(folder, recursive = TRUE)
```

The generic execution path materializes lazy output before checking it and passes
that same snapshot to the writer. This may require memory proportional to the
output. RDS serializes cooperating local writers and preserves existing versions;
it does not provide a transaction across multiple tables. See the guarantees
article before choosing a storage backend.

## Diagnose and extend

`dataraft.core::dr_plan()` inspects the definition. `dataraft.core::dr_validate(product)` checks configuration;
it does not read the delivery. After execution, `dr_quality_report(result)` shows
the evidence and `dataraft.core::dr_quality_errors(result)` returns retained local errors. After
a thrown error, use `dr_last_failure()` to retrieve the result without reading the
source again. Exported reports omit raw exception text.

Use `dataraft.core::dr_contract_policy()` for empty/extra-column and freshness policy. Use
`dataraft.core::dr_model_product()` with a `dm` for related tables. Catalog sinks are supplied by
`dataraft.adapters`. dbt, metrics and IDE integrations are optional module APIs.

The metapackage exposes 17 product verbs and `dr_demo()`. Specialist functions
are accessed through their component namespaces. Compose sources, contracts and
recipes directly on a product. Use `dr_run(x, write = FALSE)` to check it before
writing; set `stop_on_failure = FALSE` explicitly to inspect a blocked result.
