# Governance, Policies, Ports, and SLAs {#sec-governance}

Governance becomes useful when it changes a decision at a product boundary. DataRaft implements several deliberately bounded controls rather than claiming to be a general enterprise policy engine.

## Require evidence before publication

```{r}
library(dataraft)

contract <- dr_contract(
  "policies",
  columns = c(
    policy_id = "integer",
    premium = "numeric"
  ),
  key = "policy_id"
) |>
  dataraft.core::dr_contract_meta(
    owner = "Risk Analytics",
    grain = "One row per policy"
  )

product <- dr_product("policies") |>
  dr_add_contract(contract)

owner_policy <- dataraft.core::dr_policy(
  "require-owner",
  when = "publish",
  require = "owner",
  action = "block"
)

governed <- product |>
  dataraft.core::dr_add_policy(owner_policy)

dataraft.core::dr_check_policies(
  governed,
  event = "publish"
)
```

The same mechanism can warn instead of block and can target selected classifications. Organization-wide policies can also be supplied through the `dataraft.policies` option.

This is a local product policy mechanism. It is not IAM, legal approval, or a replacement for a general authorization engine.

## Ports describe product interfaces

`dr_input()` and `dr_output()` describe consumer-facing interfaces around a product.

An output port can carry a contract version, descriptive access classification, and an SLA. The `access` field is metadata, not an authorization grant.

Multiple outputs are written sequentially. They are explicitly not one distributed transaction. If a later destination fails after an earlier one commits, the result records partial publication and `dr_retry_ports()` can resume unpublished secondary outputs from the retained in-memory result.

Honest partial state is stronger than pretending heterogeneous systems share atomicity they do not provide.

## SLA evidence is about delivery

The current SLA model supports daily and weekly delivery expectations.

```{r}
daily_sla <- dataraft.core::dr_sla(
  freshness = 24,
  refresh = "daily",
  available_by = "09:00",
  timezone = "UTC"
)

delivered <- as.POSIXct(
  "2026-10-02 08:30:00",
  tz = "UTC"
)

dataraft.core::dr_check_sla(
  daily_sla,
  business_date = "2026-10-02",
  delivered_at = delivered,
  at = delivered
)
```

A quality check can pass while the delivery is late. That is why quality status and SLA status remain different evidence.

When an SLA is attached to an output port, publication requires an explicit `business_date` and records the evaluation in run metadata.

## Durable lifecycle belongs to the lake

`dataraft.lake` adds persisted lifecycle operations such as promotion, deprecation, retirement, and backfill. A retired product cannot accept new lake releases.

This creates a layered governance model:

| Layer | Responsibility |
|---|---|
| contract and quality | acceptance expectations |
| policy | bounded organization rule at a lifecycle event |
| port | declared consumer interface |
| SLA | delivery promise and evidence |
| lake lifecycle | persisted release and product state |
| external governance | IAM, legal controls, retention, approvals |

The layers can work together without pretending one mechanism owns all governance.

## Exercises

1. Remove the owner metadata from the executable example and inspect the policy decision.
2. Explain what should happen operationally if a primary output commits and a secondary output fails.
3. Evaluate the SLA with a delivery after 09:00. Why is that different from a failed quality rule?
